Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Latest
Pass over chapter 12

\autoref{fig:queueMS} shows another
concurrent queue implementation~\cite{MS96}.
It is well-known, but what is not often realized is that
it requires sequentially consistent memory, which is not
said explicitly in the paper.
As a result, the algorithm is must be coded very carefully to work
correctly with modern programming languages and
computer hardware.
The implementation uses separate
locks for the head and the tail,
allowing a <{put}> and a <{get}> operation
to proceed concurrently.  To avoid contention between the head and the tail,
the queue uses a dummy node at the head of the linked list.
Except initially, the dummy node is the last node that was dequeued.
Note that neither the <{head}> nor <{tail}> pointer are
ever <{None}>.
The problem is when the queue is empty and there are concurrent
<{get}> and <{put}> operations.
They obtain separate locks and then concurrently access the
<{next}> field in the dummy node---a data race with
undefined semantics in most environments.

@@

@@

\begin{figure}
\begin{code}
<{:input:../code/intsettest.hny}>
\end{code}
\caption{\harmonylink{code/intsettest.hny} Test code for \autoref{fig:linkedlist}}
\label{fig:intsettest}
\end{figure}

@@

\autoref{fig:setobj} gives the specification of a concurrent set object.
\texttt{SetObject()} returns a pointer to a variable that
contains an empty set, rather than returning an empty set.
Values can be added using \texttt{insert()} or deleted using \texttt{remove()}.
Method \texttt{contains()} checks if a particular value is in the list.

\autoref{fig:linkedlist} implements a concurrent set of integers using
an ordered linked list of integers without duplicates.
(\autoref{fig:intsettest} contains test code.)
The list has two dummy ``book-end'' nodes with values \texttt{-inf} and
\texttt{inf} (similar to the Python \texttt{math.inf} constant).
An invariant of the algorithm is that at any point in time the
list is ``valid,'' starting with a \texttt{-inf} node and ending
with a \texttt{inf} node.

\index{hand-over-hand locking}%
Each node has a lock, a value, and \textit{next}, a pointer to the next node
(which is <{None}> for the final \texttt{inf} node).
The \texttt{\_find(\textit{lst}, <{v}>)} helper method first
finds and locks two consecutive nodes <{before}> and \textit{after}
such that <{before->data.value < v <= after->data.value}>.
It does so by performing something called \emph{hand-over-hand locking}.
It first locks the first node, which is the \texttt{-inf} node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a tree
hand-over-hand.
Using \texttt{\_find} the <{insert}>, \texttt{remove}, and \texttt{contains}
methods are fairly straightforward.

\glsadd{linearizable}%
Like the queue in \autoref{fig:queues}(b), the implementation of the list is
\emph{linearizable}~\cite{HW90},
\index{linearizable}%
a strong notion of consistency that
makes it appear as if each of the operations executes atomically at
some point between their invocation and return.
Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization point for the <{insert}> operation coincides exactly with the update
of the <{before.next}> pointer.
The linearization point of a \texttt{contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
\texttt{inf} node.

\section*{Exercises}
\begin{problems}
\item Add methods to the data structure in \autoref{fig:linkedlist}
that report the size of the list, the minimum value in the
list, the maximum value in the list, and the sum of the values in the list.
Are they linearizable?  If so, what are their linearization points?
\item Create a thread-safe sorted binary tree.  Implement a module \texttt{bintree}
with methods
$\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts <{v}> into
tree <{t}>, and $\mathtt{contains}(t, v)$ that checks if <{v}> is in tree <{t}>.  Use a single
lock per binary tree.
\item Create a binary tree that uses, instead of a single lock per tree,
a lock for each node in the tree.
\end{problems}

>>>>

\autoref{fig:queueMS} shows another
concurrent queue implementation~\cite{MS96}.
It is well-known, but what is not often realized is that
it requires sequentially consistent memory, which is not
said explicitly in the paper.
As a result, the algorithm must be coded very carefully to work
correctly with modern programming languages and
computer hardware.
The implementation uses separate
locks for the head and the tail,
allowing a <{put}> and a <{get}> operation
to proceed concurrently.  To avoid contention between the head and the tail,
the queue uses a dummy node at the head of the linked list.
Except initially, the dummy node is the last node that was dequeued.
Note that neither the <{head}> nor <{tail}> pointer are
ever <{None}>.
The problem is when the queue is empty and there are concurrent
<{get}> and <{put}> operations.
They obtain separate locks and then concurrently access the
<{next}> field in the dummy node---a data race with
undefined semantics in most environments.

@@

\begin{figure}
\begin{code}
<{:input:../code/intsettest.hny}>
\end{code}
\caption{\harmonylink{code/intsettest.hny} Test code for set objects}
\label{fig:intsettest}
\end{figure}

@@

@@

\autoref{fig:setobj} gives the specification of a concurrent set object.
\texttt{SetObject()} returns a pointer to a variable that
contains an empty set, rather than returning an empty set \emph{value}.
As such, it is more like an object in an object-oriented language than
like a value in its own right.
Values can be added to the set object using
<{insert()}> or deleted using <{remove()}>.
Method <{contains()}> checks if a particular value is in the list.
\autoref{fig:intsettest} contains a simple (although not very thorough)
test program to demonstrate the use of set objects.

\autoref{fig:linkedlist} implements a concurrent set object using
an ordered linked list of integers without duplicates.
The list has two dummy ``book-end'' nodes with values <{-inf}> and
<{inf}> (similar to the Python \texttt{math.inf} constant).
An invariant of the algorithm is that at any point in time the
list is ``valid,'' starting with a <{-inf}> node and ending
with an <{inf}> node.

\index{hand-over-hand locking}%
Each node has a lock, a value, and <{next}>, a pointer to the next node
(which is <{None}> for the <{inf}> node).
The <{_find(lst, v)}> helper method first
finds and locks two consecutive nodes <{before}> and <{after}>
such that <{before->data.value < v <= after->data.value}>.
It does so by performing something called \emph{hand-over-hand locking}.
It first locks the first node, which is the <{-inf}> node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a rope
hand-over-hand.
Using \texttt{\_find}, the <{insert}>, \texttt{remove}, and \texttt{contains}
methods are fairly straightforward to implement.

\begin{comment}
\glsadd{linearizable}%
Like the queue in \autoref{fig:queues}(b), the implementation of the list is
\emph{linearizable}~\cite{HW90},
\index{linearizable}%
a strong notion of consistency that
makes it appear as if each of the operations executes atomically at
some point between their invocation and return.
Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization point for the <{insert}> operation coincides exactly with the update
of the <{before.next}> pointer.
The linearization point of a \texttt{contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
<{inf}> node.
\end{comment}

\section*{Exercises}
\begin{problems}
\item Add methods to the data structure in \autoref{fig:linkedlist}
that report the size of the list, the minimum value in the
list, the maximum value in the list, and the sum of the values in the list.
(All these should ignore the two end nodes.)
% Are they linearizable?  If so, what are their linearization points?
\item Create a thread-safe sorted binary tree.  Implement a module \texttt{bintree}
with methods
$\mathtt{BinTree}()$ to create a new binary tree, $\mathtt{insert}(t, v)$ that inserts <{v}> into
tree <{t}>, and $\mathtt{contains}(t, v)$ that checks if <{v}> is in tree <{t}>.  Use a single
lock per binary tree.
\item Create a binary tree that uses, instead of a single lock per tree,
a lock for each node in the tree.
\end{problems}
